---
title: "Assignment 1"
author: "Nam Bui, 300665058"
date: "01/08/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(error=TRUE)
```



# Q1.

### a. Number of adults living in the same household  
- Variable type: Numerical, Integer  
- R class: integer  
- R type: integer  
    
### b. Preferred political party  
- Variable type: Categorical, Nominal  
- R class: factor / character  
- R type: integer (if factor) / character (if character)  

### c. Whether an individual voted in the most recent election  
- Variable type: Categorical, Boolean / Numerical, Binary
- R class: logical (if Boolean) / numeric (if Binary)  
- R type: logical (if Boolean) / integer (if Binary)  

### d. Water temperature  
- Variable type: Numerical, Continuous  
- R class: numeric  
- R type: double  

### e. Self-reported level of perceived stress, from none to very high  
- Variable type: Categorical, Ordinal  
- R class: ordered factor / character  
- R type: integer (if ordered factor) / character (if character)  
    
    
    
# Q2.

```{r}
y <- c("Fred" = 5,"Wilma" = 1,"Betty" = 8,"Barney" = 3)
```
### a. 
```{r}
# Get elements in y with values greater than 3
y[y>3]
```
### b.
```{r}
# Get the names of the elements in y with values equal to 3
names(y[y==3])
```
### c.
```{r}
# Get the positions of the elements with maximum values in y
which(y==max(y))
```
### d.
```{r}
a <- c(5,2,3) # Simple vector
z <- array(c(2,0,5,1,1,2), dim=c(3,2)) # 2D 3x2 array 
a
z
```
### e.
```{r}
# t(z) returns the transposition of matrix z
tz <- t(z)
tz
```
### f.
R computes `2*z` (z is presumably the above 3x2 matrix) by applying the "times 2" mathematical operation on each of the elements in z, and use those results to construct a matrix of the same dimensions.



# Q3.

### a.
```{r}
dot_product <- function(x,y) {
  valid <- FALSE
  result <- 0
  
  # Loop through all positions of vector x (presuming x and y are the same length)
  for (i in 1:length(x)) { 
    if (!is.na(x[i]) && !is.na(y[i])) {
      
      # If both positions are not NA, add their product to the result
      result <- result + x[i]*y[i] 
      valid <- TRUE
    }
  }
  
  # If still hasn't found a valid pair, print warning message 
  if (!valid) stop("No valid pairs of values to compute the product.") 
  return(result)
}
```
### b.
```{r}
a <- c(NA,2,NA)
b <- c(1,NA,5)
dot_product(a,b)
```
### c.
```{r}
a <- c(5,2,1,NA,1)
b <- c(1,NA,5,2,4)
dot_product(a,b)
```
### d.
```{r}
dot_product <- function(x,y) {
  
  # If x and y are not the same length, print warning message
  if (length(x)!=length(y)) stop("Vectors must be the same length.")
  valid <- FALSE
  result <- 0
  
  # Loop through all positions of vector x (presuming x and y are the same length)
  for (i in 1:length(x)) { 
    if (!is.na(x[i]) && !is.na(y[i])) {
      
      # If both positions are not NA, add their product to the result
      result <- result + x[i]*y[i] 
      valid <- TRUE
    }
  }
  
  # If still hasn't found a valid pair, print warning message 
  if (!valid) stop("No valid pairs of values to compute the product.") 
  return(result)
}
```
### e.
```{r}
c <- c(1,NA,5,2)
dot_product(a,c)
```



# Q4.

### a.
```{r}
dinos <- read.csv("dinosaurs.csv")
```
### b.
```{r}
dim(dinos)
```
### c.
```{r}
dinos[1:4,]
```
### d.
```{r}
# Replace empty strings in diet with NA
dinos$diet[dinos$diet==""] <- NA
# Count number of NA occurrences after replacement
sum(is.na(dinos$diet))
```
### e.
```{r}
is.factor(dinos$diet)
diet_factor <- factor(dinos$diet, levels=c("herbivorous", "carnivorous", "omnivorous"))
```
### f.
```{r}
boxplot(dinos$length_m ~ diet_factor,
        main = "Length by Diet Type (in meters)", # title
        xlab = "Diet", # x-axis label
        ylab = "Length (meters)") # y-axis label
```



# Q5.
`x=y`: This is an assignment statement that assigns a value of y to x. While it can work for general assignments, it is discouraged in R style, and is preferably used for assigning values to function arguments.
`x==y`: This is an element-wise comparison that returns either TRUE if the element of x is equal to the corresponding element of y, or FALSE otherwise, for each pair of elements.   
`x<-y`: This is an assignment statement that assigns a value of y to x, and is the preferred assignment operator in R code used to assign values to variables.
`x<(-y)`: This is an element-wise comparison that returns either TRUE if the element of x is less than the negative of the corresponding element of y, or FALSE otherwise, for each pair of elements.  
`x^(-y)`: This is a element-wise power operation that returns a vector where each element of x is raised to the power of the negative of the corresponding element of y



# Q6.
```{r}
library(dplyr)
data("starwars")
```
### a. 
```{r}
hist(starwars$height,
     main = "Histogram of Star Wars Characters' Height",
     xlab = "Height (cm)")
```
### b.
```{r}
plot(starwars$height, starwars$mass,
     main = "Scatterplot of Height vs Mass",
     xlab = "Height (cm)",
     ylab = "Weight (kg)")
```
Comment: Generally, taller characters tend to be heavier, except for one instance whose weight is disproportionately high relative to their height.
### c.
```{r}
boxplot(starwars$height ~ starwars$gender,
        main = "Boxplot of Star Wars Characters' Height by Gender",
        xlab = "Gender",
        ylab = "Height (cm)",
        col = "blue")
```
### d.
```{r}
Num_homeworld <- table(starwars$homeworld)
Cat1entri <- names(Num_homeworld)[Num_homeworld == 1]
starwars$homeworld2 <- ifelse(starwars$homeworld %in% Cat1entri, "Other", starwars$homeworld)

# Create a contingency table
counts <- table(starwars$gender, starwars$homeworld2)

# Draw barplot with legend
barplot(counts,
        beside = TRUE, # To make grouped bar chart instead of stacked
        legend.text = TRUE, # Enable legend
        main = "Counts of Characters by Gender and Homeworld2",
        xlab = "Homeworld",
        ylab = "Count",
        las = 2)  # Rotate x-axis labels for readability
```

# Q7.

### Error 1:
```{r eval=FALSE}
# Define x
x <- matrix(c(3,0,5,6,NA 3), ncol = 2)
```
- Error: Missing comma between `NA` and `3`
- Fix: Add a comma between `NA` and `3`
```{r}
x <- matrix(c(3,0,5,6,NA,3), ncol = 2)
x
```
### Error 2:
```{r eval=FALSE}
# Store the second column of x as a new object called y
y <- x[2]
```
- Error: `x[2]` only retrieves the second element from the first column, not the entire second column
- Fix: use `x[,2]` (the blank before the comma indicates that the operation should cover all rows the the second column)
```{r}
y <- x[,2]
y
```
### Error 3:
```{r eval=FALSE}
# Calculate the mean value of y after excluding NA entries
mean_y <- mean(y, na.remove=TRUE)
```
- Error: `na.remove` is not a valid argument for the `mean()` function
- Fix: use `na.rm` instead
```{r}
mean_y <- mean(y, na.rm=TRUE)
mean_y
```
### Error 4:
```{r eval=FALSE}
# Fetch the entries of y that are bigger or equal to 2
y[y => 2]
```
- Error: The `=>` is not a valid conditional operator
- Fix: Use the "greater than or equal to" conditional operator `>=`
```{r}
y[y >= 2]
```


